Towards an Ecological Trait-data Standard

### 2018-05-02 manuscript/manuscript.Rmd
finalize ms v0.7, shared with coauthors

---
title: "Towards an Ecological Trait-data Standard"
author:
- name: Florian D. Schneider \footnotemark[1] \footnotemark[2]
- name: Malte Jochum \footnotemark[4]
- name: Gaetane Le Provost \footnotemark[2]
- name: Andreas Ostrowski \footnotemark[5]
- name: Caterina Penone \footnotemark[4]
- name: Markus Fischer \footnotemark[4]\footnotemark[2]
- name: David Fichtmüller \footnotemark[5]
- name: Anton Güntsch \footnotemark[5]
- name: Pete Manning \footnotemark[2]
- name: Birgitta König-Ries\footnotemark[6]
- name: Nadja K. Simons \footnotemark[7]
csl: amnat.csl
fontsize: 12pt
geometry: margin=1in
output:
  pdf_document:
    fig_caption: yes
    fig_crop: no
    keep_tex: yes
    template: pretty.template
  html_document:
    fig_caption: yes
    fig_width: 4
    highlight: null
    keep_md: yes
    theme: united
  word_document:
    fig_caption: yes
    fig_width: 6
bibliography: div05.bib
---

@@

\footnotetext[1]{Corresponding author: florian.dirk.schneider@gmail.com}
\footnotetext[2]{Senckenberg Biodiversity and Climate Research Centre (BiK-F), Senckenberganlage 25, 60325 Frankfurt am Main, Germany}
\footnotetext[4]{Institute of Plant Sciences, University of Bern, Bern, Switzerland}
\footnotetext[5]{Botanischer Garten und Botanisches Museum Berlin-Dahlem, Freie Universität Berlin, Berlin, Germany}
\footnotetext[6]{Department of Mathematics and Computer Science, Friedrich-Schiller-Universität Jena, Jena, Germany}
\footnotetext[7]{Terrestrial Ecology Research Group, Department for Ecology and Ecosystem Management, Center for Life and Food Sciences Weihenstephan, Technische Universität München, Freising, Germany}

@@

functional ecology, functional traits, semantic web, ontologies, data standard

@@

Functional traits are phenotypic characteristics that are related to the fitness and performance of an organism [@mcgill06; @violle07]. Because trait-based approaches to ecological and evolutionary research questions allow for both mechanistic insight (refs) and generality (refs) recent years have seen a proliferation of trait-based research in a wide range of fields.
Trait-based studies have spanned the evolutionary basis of individual-level properties (refs) to global patterns of biodiversity [@diaz16] and ecosystem functioning (refs). 
For example, using traits is a promising venue to bypass taxonomic impediment, i.e. the fact that a majority of species are yet undescribed and little is known of their interactions with the environment and other organisms (refs), since they allow inferring the ecological role of organisms [@duarte11; @schrodt15] from their apparent features.

@@

<!-- 3. trait databases have been developed, with standard structure and terms --> 
To this end, trait data have been standardized in centralized databases for specific organism groups and regional scope, often centered around specific research questions (ref to TRY, panthera, etc.). 
These compiled data are either being published as static data tables that are the product of a one-time effort of collecting and harmonizing data (ref Amphibio, panthera, mammals), or in the form of sophisticated online portals, which attract data submissions from a focal community of researchers and provide research networks for consensus building in data resolution and measurement methods [e.g. @kattge11]. Furthermore, these portals map heterogeneous data into a common scheme, and importantly also, offer access control and data usage policies, to protect the rights of the original data providers, while simplifying data queries for synthesis researchers who seek to utilize these often-complex datasets. 
To enable the compatibility of data across data-bases, semantic-web standards [@wieczorek12; @page08] and ontologies of standard terms [@walls12; @garnier17] for trait-data are beeing developed.
Solving all these issues of data assemblage and harmonization in centralised infrastructures requires a huge effort in costs and personell (see below for a review of current initiatives).

@@

# 1. A review of trait-standard initiatives

In this section, we review four kinds of initiatives that are of relevance for trait data standardisation: 
- The first are projects that work with trait datasets which have been assembled out of a particular research interest, either by measurement or collated from literature; 
- the second aim to harmonize trait data from the literature or from direct measurement into trait databases and make those data widely available; 
- the third aim at the standardization and develops consensus methods and definitions for traits and provides standard terminologies; 
- the fourth kind of initiative feed the needs of biodiversity informatics and provide database structures and relational definitions to construct relational 'ontologies' for an application in a 'semantic web' of information.

@@

<!-- Fig 1 here? -->

@@

<!-- community database backend, but static publication --> 
In the past two decades,  distributed trait data have been aggregated and harmonized into greater collections with particular taxonomic or regional focus [e.g. @klotz02; @iversen17; @kleyer08; @oliveira17; @iversen17; @myhrvold15; @kissling14; @jones09]. <!--[@wright04; ] -->
While mostly concerned with issues of heterogeneity in units or factor levels, and aiming for high taxonomic coverage, few of these datasets are using a standardised terminology for taxa or for traits that would allow to relate them to other databases.<!--and aiming for taxonomic completeness, the products of these initiatives employ a project specific structure and vocabulary. -->
Documentation of metadata and methodology differs in detail, depending on the research focus of the initiative. 
Just as the individual datasets described above, many of these databases are published as static datasets on low-threshold file hosting platforms and are updated irregularly.

@@

Wherever referring to individual terms in publications or metadata, please do so via their global identifier, which is hosted by the GFBio Terminology Service [@karam14; https://terminologies.gfbio.org/], which also can be accessed programmatically (i.e. via the Application Programming Interface, API **in preparation!**). Wherever our glossary refines or duplicates existing terms from other ontologies for biological data, like the Glossary of EOL (http://eol.org/info/516) and Darwin Core (http://rs.tdwg.org/dwc/terms/), we indicate this in the fields 'refines' or 'identical', respectively.

@@

If no globally available trait definition is available that can be referenced, trait-datasets should be accompanied by a dataset-specific glossary of traits, or thesaurus. Such a 'controlled vocabulary' would, in its simplest form, assign trait names with A) an unambiguous definition of the trait and B) an expected format  of measured values or reported facts (e.g. units or legit factor levels).

By providing a minimal vocabulary for this case (https://ecologicaltraitdata.github.io/ETS/#terms-for-thesauri), we hope to facilitate the publication of trait thesauri developed for a specific project context which should always be referenced in the core trait dataset. The thesaurus might accompany the core data in a separate file (e.g. within a Darwin Core Archive), or be published on any other stable internet address (via URIs). 
<!-- PM: Is this possible in practice?- in plant traits slightly different methods makes some traits comparable conceptually but not in practice, and there are lots of very minor variations of similar traits- see the vast list of traits in TRY for example -->
See our online reference for recommendations how to publish a minimal Thesaurus compliant with semantic web standards [https://ecologicaltraitdata.github.io/ETS/thesauri.html].  We also provide an example thesaurus, which has been developed to capture traits assessed for multiple animal taxa within the German Biodiversity Exploratories project on Github (http://ecologicaltraitdata.github.io/TraitDataList/).

@@

Information about the authorship and ownership of the data and the terms of use should be considered when sharing and working with trait datasets.
We define a set of terms (https://ecologicaltraitdata.github.io/ETS/#metadata-vocabulary) that allows trait data to be related to authorship, while also stating a bibliographic reference and license model.

In the case of raw measurement data, this information applies to the entire trait dataset, and should be stored along with the published data as metadata (e.g. in a separate file or on the download website). 
In cases where individual measurements and facts from different sources are compiled into a trait database, these metadata information must be provided at the measurement level. This can be achieved by appending the information as columns to the core dataset, or via an unambiguous `datasetID` and a descriptive `datasetName`.

@@

The data standard provides two extensions of the vocabulary that should be used to describe these information:

@@

Since the intention of the package is to facilitate upload of own data to hosting services but also to simplify the harmonization of published trait data, the package offers direct access to trait datasets that have been published in the public domain or under creative commons licenses. <!--See a full list of linked datasets on the package  website or by calling `data(package = "traitdataform")`.--> We invite users and authors of datasets to add further data to the package and thereby contribute to this registry for distributed trait datasets.

@@

To date, a rich distributed body of independently published trait datasets focus on particular organism groups, ecosystem types or regions. To make these heterogeneous sources of data more accessible and interlinkable, we propose a traitdata standard terminology and formatting scheme that can be applied in multiple contexts, for instance for building an own taxon-specific database or in user-side compilations of trait data from different sources.

The tools proposed here help to standardize trait datasets before upload to central as well as distributed data repositories. In times of increasing demand for open research data and international platforms for biodiversity data management, the development of meaningful terminologies for the standardisation of biodiversity data is more than essential: defined ontologies enable researchers to relate published datasets to each other to achieve a greater synthesis, thereby paving the way for new insights on global biodiversity patterns. In terms of data science, semantic web ontologies are a prerequisite for advanced big data mining using machine learning techniques.

We therefore encourage the development of trait thesauri in ontologies. The biggest challenge in community efforts of standardization of traits may be the investment in consensus building which leads to an acceptance and establishment of the methodological and conceptual definitions of traits. This effort is time consuming and lengthy, but it returns  great scientific benefit by enabling global synthesis, as was exemplified by the consistent global patterns that were unraveled by analyses of the plant trait database TRY [@diaz16].

Furthermore, the tools provided here can be applied to harmonise and combine data from heterogeneous sources. Such tools to handle the extensive raw data resources on traits are a prerequisite for research on trait variation along environmental gradients or intra-specific variation. This also opens the ground for a development of best practice principles for the filtering of raw measurements when creating species-level mean values. 
We hope this aids research on the response of traits to environmental and anthropogenic drivers, synthesis on trait patterns on all ecological scales [@messier10], as well as multi-trophic trait analysis [@soliveres16; @allan15].

@@

FDS, MJ, GP, AO, CP, and NKS conceived the ideas for the manuscript and critically revised the vocabulary for the trait data standard; All authors contributed to writing the paper under the lead of FDS; NKS authored the list of traits; FDS developed the R package. All authors contributed critically to the drafts and gave final approval for publication.

@@

The online reference for the Ecological Traitdata Standard Vocabulary described in this paper is https://ecologicaltraitdata.github.io/ETS/

@@

The living spreadsheet on existing trait datasets and databases is found at

Any future development of the Vocabulary and the R-package is coordinated via https://github.com/EcologicalTraitData/ .

>>>>

---
title: "Towards an Ecological Trait-data Standard"
author:
- name: Florian D. Schneider \footnotemark[1] \footnotemark[2] \footnotemark[3]
- name: Malte Jochum \footnotemark[4]
- name: Gaetane Le Provost \footnotemark[3]
- name: Andreas Ostrowski \footnotemark[5]
- name: Caterina Penone \footnotemark[4]
- name: David Fichtmüller \footnotemark[5]
- name: Anton Güntsch \footnotemark[5]
- name: Pete Manning \footnotemark[3]
- name: Birgitta König-Ries\footnotemark[6]
- name: Nadja K. Simons \footnotemark[7]
csl: amnat.csl
fontsize: 12pt
geometry: margin=1.3in
output:
  word_document:
    fig_caption: yes
    fig_width: 6
  html_document:
    fig_caption: yes
    fig_width: 4
    highlight: null
    keep_md: yes
    theme: united
  pdf_document:
    fig_caption: yes
    fig_crop: no
    keep_tex: yes
    template: pretty.template
bibliography: div05.bib
---

@@

\footnotetext[1]{Corresponding author: florian.dirk.schneider@gmail.com}
\footnotetext[2]{Institute of Linguistics and Literary Studies, Technische Universität Darmstadt, Darmstadt, Germany}
\footnotetext[3]{Senckenberg Biodiversity and Climate Research Centre (BiK-F) Frankfurt am Main, Germany}
\footnotetext[4]{Institute of Plant Sciences, University of Bern, Bern, Switzerland}
\footnotetext[5]{Botanischer Garten und Botanisches Museum Berlin-Dahlem, Freie Universität Berlin, Berlin, Germany}
\footnotetext[6]{Department of Mathematics and Computer Science, Friedrich-Schiller-Universität Jena, Jena, Germany}
\footnotetext[7]{Terrestrial Ecology Research Group, Department for Ecology and Ecosystem Management, Center for Life and Food Sciences Weihenstephan, Technische Universität München, Freising, Germany}

@@

functional ecology, species traits, semantic web, ontologies, data standard

@@

Species traits are phenotypic characteristics that are related to the fitness and performance of an organism [@mcgill06; @violle07]. Because trait-based approaches to ecological and evolutionary research questions allow for both mechanistic insight (refs) and generality (refs) recent years have seen a proliferation of trait-based research in a wide range of fields.
Trait-based studies have spanned the evolutionary basis of individual-level properties (refs) to global patterns of biodiversity [@diaz16] and ecosystem functioning (refs). 
For example, using traits is a promising venue to bypass taxonomic impediment, i.e. the fact that a majority of species are yet undescribed and little is known of their interactions with the environment and other organisms (refs), since they allow inferring the ecological role of organisms [@duarte11; @schrodt15] from their apparent features.

@@

<!-- 3. trait databases have been developed, with standard structure and terms --> 
To this end, trait data have been standardized in centralized databases for specific organism groups and regional scope, often centered around specific research questions [e.g. TRY, @kattge11; PanTHERIA, @jones09; AmphiBio, @oliveira17]. 
These compiled data are either being published as static data tables that are the product of a one-time effort of collecting and harmonizing data (in the case of AmphiBIO), or in the form of sophisticated online portals, which attract data submissions from a focal community of researchers and provide research networks for consensus building in data resolution and measurement methods [e.g. TRY, @kattge11]. Furthermore, these portals map heterogeneous data into a common scheme, and importantly also, offer access control and data usage policies, to protect the rights of the original data providers, while simplifying data queries for synthesis researchers who seek to utilize these often-complex datasets. 
To enable the compatibility of data across data-bases, semantic-web standards [@wieczorek12; @page08] and ontologies of standard terms [@walls12; @garnier17] for trait-data are beeing developed.
Solving all these issues of data assemblage and harmonization in centralised infrastructures requires a huge effort in costs and personell (see below for a review of current initiatives).

@@

# 1. A review of initiatives for trait-data standardisation

In this section, we review four kinds of initiatives that are of relevance for trait data standardisation:

* The first are projects that work with trait datasets which have been assembled out of a particular research interest, either by measurement or collated from literature; 
* the second aim to harmonize trait data from the literature or from direct measurement into trait databases and make those data widely available; 
* the third aim at the standardization and develops consensus methods and definitions for traits and provides standard terminologies and 'ontologies'; 
* the fourth kind of initiative feed the needs of biodiversity informatics and develop relational database structures and provide platforms to relate trait data to a wider 'semantic web' of biodiversity data.

@@

@@

<!-- community database backend, but static publication --> 
In the past two decades,  distributed trait data have been aggregated and harmonized into greater collections with particular taxonomic or regional focus [e.g. @klotz02; @iversen17; @kleyer08; @oliveira17; @myhrvold15; @kissling14; @jones09]. <!--[@wright04; ] -->
While mostly concerned with issues of heterogeneity in units or factor levels, and aiming for high taxonomic coverage, few of these datasets are using a standardised terminology for taxa or for traits that would allow to relate them to other databases.<!--and aiming for taxonomic completeness, the products of these initiatives employ a project specific structure and vocabulary. -->
Documentation of metadata and methodology differs in detail, depending on the research focus of the initiative. 
Just as the individual datasets described above, many of these databases are published as static datasets on low-threshold file hosting platforms and are updated irregularly.

@@

Wherever referring to individual terms in publications or metadata, please do so via their global identifier, which is hosted by the GFBio Terminology Service [@karam16; https://terminologies.gfbio.org/], which also can be accessed programmatically (i.e. via the Application Programming Interface, API **in preparation!**). Wherever our glossary refines or duplicates existing terms from other ontologies for biological data, like the Glossary of EOL (http://eol.org/info/516) and Darwin Core (http://rs.tdwg.org/dwc/terms/), we indicate this in the fields 'refines' or 'identical', respectively.

@@

If no globally available trait definition is available that can be referenced, trait-datasets should be accompanied by a dataset-specific glossary of traits, or thesaurus. Such a 'controlled vocabulary' would, in its simplest form, assign trait names with an unambiguous definition of the trait and an expected format  of measured values or reported facts (e.g. units or legit factor levels).

By providing a minimal vocabulary for trait thesauri (https://ecologicaltraitdata.github.io/ETS/#terms-for-thesauri), we hope to facilitate the publication of trait thesauri developed for a specific project context, which can be referenced in the trait dataset. The thesaurus might accompany the core data in a separate file (e.g. within a Darwin Core Archive), or be published on any other stable internet address (ideally providing stable URIs). 
<!-- PM: Is this possible in practice?- in plant traits slightly different methods makes some traits comparable conceptually but not in practice, and there are lots of very minor variations of similar traits- see the vast list of traits in TRY for example -->
See our online reference for recommendations how to publish a minimal thesaurus compliant with semantic web standards [https://ecologicaltraitdata.github.io/ETS/thesauri.html].  We also provide an example thesaurus, which has been developed to capture traits assessed for multiple animal taxa within the German Biodiversity Exploratories project on Github (http://ecologicaltraitdata.github.io/TraitDataList/).

@@

Information about the authorship and ownership of the data and the terms of use should be considered when sharing and working with trait datasets.
We define a vocabulary (https://ecologicaltraitdata.github.io/ETS/#metadata-vocabulary) that allows trait data to be related to authors and owners, while also stating a bibliographic reference and license model.

In the case of primary measurement data, this information applies to the entire trait dataset, and should be stored along with the published data as metadata (e.g. in a separate metadata file, possibly consider applying the ecological metadata language, EML). 
In cases where individual measurements and facts from different sources are compiled into a trait database, these metadata information must be provided at the measurement level. This can be achieved by appending the information as columns to the core dataset, or via an unambiguous `datasetID` and a descriptive `datasetName`.

@@

The data standard provides three extensions of the vocabulary that should be used to describe these information:

@@

Since the intention of the package is also to simplify the harmonization of published trait data, the package offers direct access to trait datasets that have been released in the public domain or under creative commons licenses. <!--See a full list of linked datasets on the package  website or by calling `data(package = "traitdataform")`.--> We invite users and authors of datasets to add further data to the package and thereby contribute to this registry for distributed trait datasets.

@@

To date, a rich distributed body of independently published trait datasets focus on particular organism groups, ecosystem types or regions. To make these heterogeneous sources of data more accessible and interlinkable, we propose a traitdata standard terminology that can be applied in multiple contexts, for instance for building an own taxon-specific database or in user-side compilations of trait data from different sources. It also can be applied in the communication of data content between databases. In addition, we also offer a software tool to handle the extensive raw data resources on traits.

These tools are to facilitate research on trait variation along environmental gradients or intra-specific variation based on distributed data sources. This also opens the ground for a development of best practice principles for the filtering of raw measurements when creating species-level mean values.

The tools proposed here help to standardise trait datasets before upload to central as well as distributed data repositories. In times of increasing demand for open research data and international platforms for biodiversity data management, the development of meaningful terminologies for the standardisation of biodiversity data is more than essential: defined ontologies enable researchers to relate published datasets to each other to achieve a greater synthesis, thereby paving the way for new insights on global biodiversity patterns. In terms of data science, semantic web ontologies are a prerequisite for advanced big data mining using machine learning techniques.
We hope this aids research on the response of traits to environmental and anthropogenic drivers, synthesis on trait patterns on all ecological scales [@messier10], as well as multi-trophic trait analysis.

We also encourage the development of trait thesauri in more advanced and interrelated ontologies. The biggest challenge in community efforts of standardization of traits may be the investment in consensus building which leads to an acceptance and establishment of the methodological and conceptual definitions of traits. This effort is time consuming and lengthy, but it returns  great scientific benefit by enabling global synthesis, as was exemplified by the consistent global patterns that were unraveled by analyses of the plant trait database TRY [@diaz16].

@@

FDS, MJ, GP, AO, CP, and NKS conceived the ideas for the manuscript and critically revised the vocabulary for the trait data standard; All authors contributed to writing the paper under the lead of FDS; NKS authored the list of traits; FDS developed the R package 'traitdataform'. All authors contributed critically to the drafts and gave final approval for publication.

@@

The online reference for the Ecological Traitdata Standard Vocabulary described in this paper is https://ecologicaltraitdata.github.io/ETS/, stable DOI representing all versions: 10.5281/zenodo.1041732.

Any future development of the Vocabulary and the R-package is coordinated via https://github.com/EcologicalTraitData/.

@@

The living spreadsheet on existing trait datasets and databases is found at https://goo.gl/QxzfHy.
